Exploratory data structure comparisons: Three new visual tools based on Principal Component Analysis

### 2019-12-19 article/Old versions/JAS/PCADSC_article_JAS_resub_jan2020.tex

\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}

@@

@@

We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences, namely the \textit{cumulative eigenvalue (CE) plot}, the \textit{angle plot}, and the \textit{chroma plot}. These plots are dsecribed in turn below. For the deepest understanding of the data structure differences in two datasets, we suggest using all of the three plots in the same order as they are presented. While we describe the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle}, and \ref{plot.simChroma} for the CE-, angle-, and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}

@@

Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed. They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences.

@@

\section{ESS psychological well-being scale definitions}
\label{appendix.essscales}
\begin{table}[H]
\renewcommand\thetable{C1}
\centering
\caption{ESS psychological well-being scale definitions. Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}. Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
\label{table:items}
\scriptsize
\rowcolors{1}{}{lightgray}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{ll}
\hline
Scale & Items \\
\hline
\rowcolor{lightgray} & How satisfied with life as a whole \\
\rowcolor{lightgray}\multirow{-2}{*}{Evaluative well-being }& How happy are you\\

@@

>>>>

\newcommand{\RR}{\mathbb{R}}
\DeclareMathOperator*{\V}{V}
\DeclareMathOperator*{\argmax}{argmax}
\newcommand{\R}[1]{\texttt{#1}}
\newcommand{\acos}{\text{arccos}}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}

@@

\let\thefootnote\relax\footnote{Note: An early draft of this article was included in the conference proceedings \textit{Symposium i Anvendt Statistik 2018. Copenhagen: The National Research Center for the Working Environment in Denmark \& University of Copenhagen (2018)}.}

@@

We present three diagnostic plots that are designed to shine a light on different types and levels of data structure differences, namely the \textit{cumulative eigenvalue (CE) plot}, the \textit{angle plot}, and the \textit{chroma plot}. These plots are described in turn below. For the deepest understanding of the data structure differences in two datasets, we suggest using all of the three plots in the same order as they are presented. While we describe the three plot types from a purely theoretical point of view below, we also provide example illustrations as a supplement to the somewhat technical definitions. These plots are available in Figures \ref{plot.simCE}, \ref{plot.simAngle}, and \ref{plot.simChroma} for the CE-, angle-, and chroma plots, respectively, and they are based on two simulated datasets:
\begin{description}
\item[Dataset A:] This dataset contains 1000 independent simulations from the same underlying model, namely \textit{model 1} from Figure~\ref{fig:simGraphs}. The observations are randomly divided into two groups.
\item[Dataset B:] This dataset contains 500 independent simulations from \textit{model 1} from Figure~\ref{fig:simGraphs} and 500 independent simulations from \textit{model 2} in the same figure. The observations are naturally divided into groups corresponding to which model they were simulated from.
\end{description}

@@

Three new tools, referred to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC), for the task of deciding if two datasets can be combined for analysis were proposed and discussed. They all employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create three intuitive visualizations of data structure differences. \blue{The first visualization, the cumulative eigenvalue plot, compares the eigenvalues from the two datasets. The second visualization, the angle plot, uses the interpretation of eigenvalues as explained variance to focus a comparison of loadings (eigenvectors). Finally, the third visulization, the chroma plot, allows for qualitative explanations of differences in loading patterns in selected principal components.}

@@

\section{ESS psychological well-being scale definitions}
\label{appendix.essscales}
\begin{table}[H]
\renewcommand\thetable{B1}
\centering
\caption{ESS psychological well-being scale definitions. Relationship between questionnaire items and scales, as defined in \cite{ESStopline5}. Before we construct the scale scores as item means, we transform the individual item scores such that they are all on a scale from 0 to 10 and such that 10 always corresponds being the most happy.}
\label{table:items}
\scriptsize
\rowcolors{1}{}{lightgray}
\bgroup
\def\arraystretch{1.5}
\begin{tabular}{ll}
\hline
Scale & Items \\
\hline
\rowcolor{lightgray} & How satisfied with life as a whole \\
\rowcolor{lightgray}\multirow{-2}{*}{Evaluative well-being }& How happy are you\\

@@

\section{ESS data structure comparisons using pairwise complete cases}
\label{appendix.pairwisecc}

This appendix presents the results of the PCADSC procedures described in Section \ref{sec:dataexample} when allowing for pairwise complete cases to contribute to the computations. The numbers of pairwise complete cases in the Danish, Bulgarian and Swedish samples, respectively, are presented in Tables \ref{tableNPairDK}, \ref{tableNPairBG} and \ref{tableNPairSE} below. We use the following abbreviations for variable names:
\begin{description}
\item[Eval.:] Evaluative well-being
\item[Emot.:] Emotional well-being
\item[Func.:] Functioning
\item[Vita.:] Vitality
\item[Comm.:] Community well-being
\item[Supp.:] Supportive relationships
\end{description}
Figures \ref{plotBG.cehair.pairwiseCC} and \ref{plotBG.pancake.pairwiseCC} show the resulting angle plot, CE plot and chroma plot for the comparison of the Danish and Bulgarian data. We see virtually no differences relative to the figures presented in Section \ref{sec:dataexample} that used complete cases only. Figures \ref{plotSE.cehair.pairwiseCC} and \ref{plotSE.pancake.pairwiseCC}  show the PCADSC plots for the comparison of Danish and Swedish data based on pairwise complete cases, and again, we find only very minor differences from the plots in Section \ref{sec:dataexample}. In conclusion, in this specific data example, the results are robust towards whether only complete cases or pairwise complete cases are used.

\begin{table}[ht]
\renewcommand\thetable{C1}
\centering
\caption{Number of observations in the Danish sample that are complete for each pairwise combination of variables. The total number of observations, including incomplete observations, is $n_\text{DK, total} = 1650$} 
\label{tableNPairDK}
\begin{tabular}{rrrrrrr}
  \hline
 & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
  \hline
Eval. & 1644 & 1608 & 1572 & 1629 & 1589 & 1613 \\ 
  Emot. & 1608 & 1613 & 1553 & 1606 & 1564 & 1588 \\ 
  Func. & 1572 & 1553 & 1573 & 1568 & 1535 & 1556 \\ 
  Vita. & 1629 & 1606 & 1568 & 1635 & 1583 & 1608 \\ 
  Comm. & 1589 & 1564 & 1535 & 1583 & 1592 & 1565 \\ 
  Supp. & 1613 & 1588 & 1556 & 1608 & 1565 & 1617 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\renewcommand\thetable{C2}
\centering
\caption{Number of observations in the Bulgarian sample that are complete for each pairwise combination of variables. The total number of observations, including incomplete observations, is $n_\text{BG, total} = 2260$} 
\label{tableNPairBG}
\begin{tabular}{rrrrrrr}
  \hline
 & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
  \hline
Eval. & 2223 & 2144 & 1975 & 2094 & 2124 & 2127 \\ 
  Emot. & 2144 & 2169 & 1960 & 2079 & 2080 & 2094 \\ 
  Func. & 1975 & 1960 & 1996 & 1921 & 1938 & 1941 \\ 
  Vita. & 2094 & 2079 & 1921 & 2119 & 2034 & 2052 \\ 
  Comm. & 2124 & 2080 & 1938 & 2034 & 2151 & 2073 \\ 
  Supp. & 2127 & 2094 & 1941 & 2052 & 2073 & 2155 \\ 
   \hline
\end{tabular}
\end{table}

\begin{table}[ht]
\renewcommand\thetable{C3}
\centering
\caption{Number of observations in the Swedish sample that are complete for each pairwise combination of variables. The total number of observations, including incomplete observations, is $n_\text{SE, total} = 1847$} 
\label{tableNPairSE}
\begin{tabular}{rrrrrrr}
  \hline
 & Eval. & Emot. & Func. & Vita. & Comm. & Supp. \\ 
  \hline
Eval. & 1843 & 1832 & 1784 & 1837 & 1801 & 1831 \\ 
  Emot. & 1832 & 1835 & 1782 & 1830 & 1795 & 1826 \\ 
  Func. & 1784 & 1782 & 1788 & 1784 & 1752 & 1780 \\ 
  Vita. & 1837 & 1830 & 1784 & 1841 & 1800 & 1830 \\ 
  Comm. & 1801 & 1795 & 1752 & 1800 & 1804 & 1794 \\ 
  Supp. & 1831 & 1826 & 1780 & 1830 & 1794 & 1835 \\ 
   \hline
\end{tabular}
\end{table}

\begin{figure}[H]
\renewcommand\thefigure{C1}
\center
\includegraphics[scale=0.8]{Figure7_PCADSC_v3_partialcomplete.pdf}
\caption{CE plot (top panel) and angle plot (bottom panel) comparing Danish and Bulgarian data using pairwise complete observations.}
\label{plotBG.cehair.pairwiseCC}
\end{figure}

\begin{figure}[H]
\renewcommand\thefigure{C2}
\center
\includegraphics[scale=0.8]{Figure8_PCADSC_v3_partialcomplete.pdf}
\caption{Chroma plot for comparing Danish and Bulgarian data  data using pairwise complete observations.}
\label{plotBG.pancake.pairwiseCC}
\end{figure}

\begin{figure}[H]
\renewcommand\thefigure{C3}
\center
\includegraphics[scale=0.8]{Figure9_PCADSC_v3_partialcomplete.pdf}
\caption{CE plot (top panel) and angle plot (bottom panel) comparing Danish and Swedish data using pairwise complete observations.}
\label{plotSE.cehair.pairwiseCC}
\end{figure}

\begin{figure}[H]
\renewcommand\thefigure{C4}
\center
\includegraphics[scale=0.8]{Figure10_PCADSC_v3_partialcomplete.pdf}
\caption{Chroma plot for comparing Danish and Swedish  data using pairwise complete observations.}
\label{plotSE.pancake.pairwiseCC}
\end{figure}
